Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-04-28 hazelnut-popl17.tex
new abstract

\authorinfo{\vspace{-2px}}{}{}
%\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Programs are rich inductive structures, but programmers typically construct and manipulate them indirectly, through flat textual representations. This indirection  comes at a cost: programmers must understand the  subtleties of parsing, and it can require several text editor actions to make a single syntactically and semantically well-defined change. 
During these sequences of edit actions, or when the programmer makes a mistake, programmers and programming tools must contend with malformed text or semantically ill-defined programs.

\emph{Structure editors} promise to alleviate these burdens by exposing only edit actions that  produce sensible changes to the program structure.
Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically ill-defined as they are being constructed.
In this paper, we report on our ongoing efforts to develop Hazelnut, a minimal  structure editor defined in a principled type-theoretic style where all edit actions leave the program in both a syntactically and semantically well-defined state. Uniquely, Hazelnut does not force the programmer to construct the program in a strictly ``outside-in'' fashion.
Formally, Hazelnut is a bidirectionally typed lambda calculus extended with
1) \emph{holes} (which mark subterms that are being  constructed from the inside out);
2) a \emph{focus model}; and 3) a bidirectional \emph{action model} equipped with a useful \emph{action sensibility} theorem.
\end{abstract}

@@

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single cursor.}
\label{fig:zexp-syntax}
\end{figure}

>>>>

%\authorinfo{\vspace{-2px}}{}{}
\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Programmers typically construct and manipulate well-typed programs indirectly, as text that must pass through a parser and typechecker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer makes a mistake, is commonly found to be malformed or ill-typed. Contending with (i.e. reasoning about) malformed text or well-formed but meaningless syntax trees is a difficult task, for programmers and  tools alike.

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.

We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the program in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a foundation for \emph{live programming environments}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

@@

\end{subequations}
\subsection{Focus Model}\label{sec:cursors}
\begin{figure}[t]
\hspace{-3px}$\arraycolsep=3pt\begin{array}{lllllll}
\mathsf{ZTyp} & \ztau & ::= &
  %\zlsel{\htau} ~\vert~
  \zwsel{\htau} ~\vert~
  %\zrsel{\htau} ~\vert~
  \tarr{\ztau}{\htau} ~\vert~
  \tarr{\htau}{\ztau} \\
\mathsf{ZExp} & \zexp & ::= &
  %\zlsel{\hexp} ~\vert~
  \zwsel{\hexp} ~\vert~
  %\zrsel{\hexp} ~\vert~
  \zexp : \htau ~\vert~
  \hexp : \ztau ~\vert~
  \hlam{x}{\zexp} ~\vert~
  \hap{\zexp}{\hexp} ~\vert~
  \hap{\hexp}{\zexp} ~\vert~
  \hadd{\zexp}{\hexp} ~\vert~
  \hadd{\hexp}{\zexp} ~\vert~
  \hhole{\zexp}
\end{array}$
%\textbf{Sort} & & & \textbf{Operational Form} & \textbf{Stylized Form} & \textbf{Description}\\
\caption{Syntax of Z-types and Z-expressions, i.e. types and expressions with holes and a single focus.}
\label{fig:zexp-syntax}
\end{figure}

### 2016-04-28 intro.tex
new abstract

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use are often presented with text that does not correspond to a well-defined program. This may be because the programmer is in the midst of a sequence of
text edit actions that leave the program ``temporarily'' malformed or ill-defined, or because the programmer has made a mistake. This complicates the programming process, both because the language definition provides no reasoning principles relevant to such text, and, relatedly, because the editor can no longer provide useful services (e.g. syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to 
Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.)

>>>>

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use are often presented with text that does not correspond to a well-defined program. This may be because the programmer is in the midst of a sequence of
text edit actions that leave the program ``temporarily'' malformed or ill-defined, or because the programmer has made a mistake. This complicates the programming process, both because the language definition provides no reasoning principles relevant to such text, and, relatedly, because the editor can no longer provide useful services (e.g. syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to

Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.)
